---
title: 'ESM 204 H3: Distributional Consequences of Climate Policy'
author: "Katie Pietrucha, Charlie Wilson, & Tommy King"
date: "5/9/2022"
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(rootSolve)
```

```{r}
# read in data
df <- read_csv("HW3_data.csv") %>% 
  select(-1) %>% 
  clean_names()
```

## Question 1
- Current electricity cost is $0.10/kWh
- One kWh electricity emits 0.85 lbs CO2
- 2204.62 lbs in 1 metric ton 
- Interim SCC value of $51/metric ton or 5100 cents/metric ton (total external cost)

```{r}
# convert lbs to metric ton
lb_to_ton_convert <- (0.85/2204.62) 
# one kWh electricity emits 0.0003856 metric tons of carbon

MEC <- 5100*lb_to_ton_convert # $.0197
``` 

- MEC = $ `r round((MEC/100), 2)` cents per kWh electricity

## Question 2

```{r}
model_demand_l <- lm(price_cents  ~ q_low_kwh, data=df)
model_demand_h <- lm(price_cents ~ q_high_kwh, data=df)

# need to rearrange the parameter to get Q(P)! 

# Qagg = Qlow(P) + Qlow(h) 

# Importantly, since their intercepts are different, we know that Qagg(P) will have a kink. I include an ifelse() statement to take care of the kink.

# define a function to get demand

demand <- function(p, model){
  q <- (p - model$coefficients[[1]])/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

# for each p level, return estimated aggregate demand
demand_agg <- function(p){
  q <- demand(p, model_demand_l) + demand(p, model_demand_h)
  return(q)
}

price = seq(0, 31.6, length.out = 100)
Qagg <- map(price, demand_agg) %>% unlist()

df2<- tibble(Qagg = Qagg, price = price)
market_q <- demand_agg(10)

ggplot() +
  geom_line(df2, mapping = aes(x = Qagg, y =price), color = "black") +# Agg demand
    geom_abline(slope = coef(model_demand_l)[["q_low_kwh"]], 
              intercept = coef(model_demand_l)[["(Intercept)"]], color="goldenrod", linetype="twodash") + # Low consumer demand
  geom_abline(slope = coef(model_demand_h)[["q_high_kwh"]], 
              intercept = coef(model_demand_h)[["(Intercept)"]], color="steelblue", linetype="twodash") + # High consumer demand
  geom_abline(slope = (10/market_q), intercept = 0, color = "green") + # supply curve
  geom_abline(slope = 0, intercept = MEC, color = "red") + # MEC
  scale_x_continuous(limits = c(0,850000), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,35), expand = c(0, 0)) +
  labs(x = "Electricity (kWh)", y = "Price (cents)") +
  theme_bw()

slope_supply <- (10/market_q)
# supply curve equation: P = 0.0000186Q

```

```{r}

# calculating surpluses at market equilibrium

# CS calculation needs Y intercept of high demand/demand agg curve --> at 0 kWh, price would be $0.316

high_sum <-summary.lm(model_demand_h)
high_y_int <- high_sum$coefficients[1, 1] 


low_sum <- summary.lm(model_demand_l)
low_y_int <- low_sum$coefficients[1, 1]

# Converted back to dollars

cs_sq_high <- 0.5*market_q*((high_y_int/100) - 0.10)
# High consumer benefit $57,984.10
cs_sq_low <- 0.5*market_q*((low_y_int/100) - 0.10)
# Low consumer benefit $35,882.29

cs_sq_total <- cs_sq_high + cs_sq_low
# Consumer benefit $93,866.39

ps_sq <- 0.5*market_q*0.10
# Producer benefit $26,835.97
env_cost_sq <- (MEC*market_q)/100
# Environmental cost $10,553.65
```

- Aggregate monthly demand curve is the black curve with a kink, which is the horizontal sum of the low (gold dashed) and high (blue dashed) demand curves.
- The supply curve is the green curve, which represents the marginal private cost curve. Its equation is P = `r round(slope_supply, 7)`Q

Under status quo:
- Consumer benefit = $`r round(cs_sq_total, 2)` 
- Producer benefit = $`r round(ps_sq, 2)`
- Environmental cost = $`r round(env_cost_sq, 2)`

## Question 3

The high-income consumers get much more of total benefit, at `r round(cs_sq_high, 2)` USD compared to `r round(cs_sq_low, 2)`USD for low-income consumers.

## Question 4

```{r}
taxcurve <- df2 - MEC # consumers are taxed by the MEC

ggplot() +
  geom_line(df2, mapping = aes(x = Qagg, y =price)) +# Agg demand
    geom_abline(slope = coef(model_demand_l)[["q_low_kwh"]], 
              intercept = coef(model_demand_l)[["(Intercept)"]], color="goldenrod", linetype="twodash") + # Low consumer demand
    geom_abline(slope = coef(model_demand_l)[["q_low_kwh"]], 
              intercept = (coef(model_demand_l)-MEC)[["(Intercept)"]], color="gray", linetype="twodash") + # new low consumer demand
  geom_abline(slope = coef(model_demand_h)[["q_high_kwh"]], 
              intercept = coef(model_demand_h)[["(Intercept)"]], color="steelblue", linetype="twodash") + # High consumer demand
  geom_abline(slope = coef(model_demand_h)[["q_high_kwh"]], 
              intercept = (coef(model_demand_h)-MEC)[["(Intercept)"]], color="gray", linetype="twodash") + # new high consumer demand
  geom_abline(slope = (10/market_q), intercept = 0, color = "green") + #supply curve equation:P= 0.0000186Q
  geom_abline(slope = 0, intercept = MEC, color = "red") + # MEC
  scale_x_continuous(limits = c(0,850000), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,35), expand = c(0, 0)) +
  labs(x = "Electricity (kWh)", y = "Price (cents)") +
  geom_line(taxcurve, mapping = aes(x = Qagg, y = price), color = "orange") + # new demand curve adjusted for tax on consumers
  theme_bw()

#find intercept of new agg demand curve and supply curve

#define supply function
supply <- function(p){
  q <- p/slope_supply
  return(q) 
}

#define demand function with tax
demand_tax <- function(p, model){
  q <- (p - (model$coefficients[[1]]-MEC))/model$coefficients[[2]]
  q <- ifelse(q<0,0,q)
  return(q)
}

demand_agg_tax <- function(p){
  q <- demand_tax(p, model_demand_l) + demand_tax(p, model_demand_h)
  return(q)
}

# set the two function equal to each other to find optimal price and quantity 
new_intercept <- uniroot(function(p)
  demand_agg_tax(p) - supply(p),
  interval = c(0,40))

p_star <- new_intercept[["root"]] # p_star is $0.09319

q_star <- supply(p_star) # q_star is 500216 kwh

#saving code below just in case we need it (old adjusted demand curve)

# new aggregate with tax 
#demand_tax <- function(p, model){
#  q <- (p - (model$coefficients[[1]]-1.97))/model$coefficients[[2]]
#  q <- ifelse(q<0,0,q)
#  return(q)
#}

#demand_agg_tax <- function(p){
#  q <- demand_tax(p, model_demand_l) + demand(p, model_demand_h) # high consumers are not subjugated to the tax
#  return(q)
#}

#price_tax = seq(0, 31.6, length.out = 100)
#Qagg_tax <- map(price, demand_agg_tax) %>% unlist()

#df2_tax<- tibble(Qagg = Qagg_tax, price = price_tax)
#market_q_tax <- demand_agg_tax(10)

```

```{r}
# calculating surpluses at taxed equilibrium

# CS calculation needs Y intercept of new demand agg curve --> at 0 kWh, price would be $0.296

tax_agg_y_int <- high_y_int - MEC

tax_low_y_int <- low_y_int - MEC

# Converted back to dollars

tax_cs_high <- 0.5*q_star*((tax_agg_y_int/100) - ((p_star/100) + MEC/100))
# High consumer benefit $45,908.74

tax_cs_low <-0.5*q_star*((tax_low_y_int/100) - ((p_star/100) + MEC/100)) - ((MEC/100)*q_star)
# Low consumer benefit $15,470.08

tax_cs_total <- tax_cs_high + tax_cs_low
# Consumer benefit $61,378.81

tax_ps <- 0.5*q_star*(p_star/100)
# Producer benefit $23,316.15
tax_env_cost <- (MEC*q_star)/100
# Environmental cost $9,837.22

# Tax revenue is the same as the environmental cost. 
```

An optimal electricity tax is equal to the MEC, $ `r round((MEC/100), 2)` cents per kWh electricity. This tax would be paid for by both low and high-income consumers, shifted both of their demand curves (indicated by gray dotted lines) and resulted in a new aggregate demand curve, the orange kinked curve. The tax would have the following effects:

(a) The electricity produced and consumed would now be `r round(q_star, 2)` kWh. Less electricity is produced and consumed.
(b) The price of electricity would be `r round(p_star/100, 4)` cents per kWh plus an additional `r round(MEC/100, 4)` cents in tax, for a total of `r round((p_star/100) + (MEC/100), 4)` cents per kWh. This signifies an overall price increase compared to the market equilibrium price of $0.10.
(c) The overall welfare of “high” income consumers would decrease to $`r round(tax_cs_high, 2)`.
(d) The overall welfare of “low” income consumers would decrease to $`r round(tax_cs_low, 2)`.
(e) The producer surplus to power suppliers (i.e., electricity producers) would decrease to $`r round(tax_ps, 2)`.
(f) The total environmental damage decreases to $`r round(tax_env_cost, 2)`.
(g) The total tax revenue generated is the same as the total environmental damage: $`r round(tax_env_cost, 2)`.

























